---
output: html_document
editor_options: 
  chunk_output_type: inline
---
# RStudio Exercise 6

```{r}
date()
```

In Exercise 6 we were asked to conduct analyses on the RATS and BPRS datasets following the techniques described in the course book, Vehkalahti and Everitt's ([2019](https://www-taylorfrancis-com.libproxy.helsinki.fi/books/mono/10.1201/9781351202275/multivariate-analysis-behavioral-sciences-kimmo-vehkalahti-brian-everitt)) *Multivariate Analysis for the Behavioral Sciences*. The trick was to swap datasets, implementing the analyses of Chapter 8 using the RATS data and Chapter 9 using the BPRS data.

### Exploring the RATS dataset

The RATS dataset contains data from a nutrition study conducted on three groups of rats ([Crowder and Hand, 1990](https://www-taylorfrancis-com.libproxy.helsinki.fi/books/mono/10.1201/9781315137421/analysis-repeated-measures-crowder-hand)). Vehkalahti and Everitt (2019: 174) summarize the study design: "the three groups were put on different diets, and each animal’s body weight (grams) was recorded repeatedly (approximately weekly, except in week seven when two recordings were taken) over a 9-week period. The question of most interest is whether the growth profiles of the three groups differ."

I begin by reading the data, converting categorical variables to factors, and confirming the structures. (I considered factorizing the time variable, but refrained from doing so as the time value is essentially continuous by nature.) Moreover, let's take a look at what the data looks like currently.

```{r}
# Reading the data. (I have already transformed the data into long form using the wrangling script.)
RATSL <- read.csv("./data/ratsl.csv", stringsAsFactors = T)
# Converting categorical variables to factors
RATSL$ID <- as.factor(RATSL$ID)
RATSL$Group <- as.factor(RATSL$Group)
# Exploring the structure of the dataset
str(RATSL)
# Viewing the data
head(RATSL)
```

On the Datacamp [course site](https://campus.datacamp.com/courses/helsinki-open-data-science/analysis-of-longitudinal-data?ex=3), the authors note that graphical displays of data are almost always useful for exposing patterns in the data, particularly when these are unexpected. Let's begin by plotting the weights of all sixteen rats, differentiating between the groups into which they were divided.

```{r}
library(dplyr); library(tidyr); library(ggplot2)
# Plotting rats
ggplot(RATSL, aes(x = Time, y = Weight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL$Weight), max(RATSL$Weight)))
```

The graphical overview makes it clear that group 1 consists of small rats (weight < 300 g). Groups 2 and 3 seem to refer to chiefly mid-size and large rats (about 400-500 g and 500-600 g). However, the division is not categorical (some overlap is visible). It is not entirely clear to me on what basis the rats have been grouped.

The graphical presentation above has the advantage of a crystal-clear grouping and the possibility to view individual progress, at least in groups 2 and 3. However, one might get an impression that the graphs are wasting visual space. An alternative form of presenting the information is given in Vehkalahti and Everitt (2019: 177), offering a quick overview of the development of the rats. A downside of this is that groups 2 and 3 are hard to separate from each other, as the linetypes are so similar. An attempt to present all groups on the same page but differentiate between groups using colours is given below. After this I will continue with the model above.

```{r}
ggplot(RATSL, aes(x = Time, y = Weight, group = interaction(Group, ID), colour = Group, linetype = Group)) +
  geom_line() +
  theme(legend.position = "top") +
  scale_y_continuous(limits = c(min(RATSL$Weight), max(RATSL$Weight))) +
  scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 20))
```

Let's explore the data further by doing a few exercises, standardising the Weight variable and plotting again with the standardised variable.

```{r}
# Standardising the Weight variable and adding a column
RATSL <- RATSL %>%
  group_by(Time) %>%
  mutate(stdWeight = ((Weight-mean(Weight))/sd(Weight))) %>%
  ungroup()
# Glimpsing the data
glimpse(RATSL)
# Plotting again with the standardised Weight variable
ggplot(RATSL, aes(x = Time, y = stdWeight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") +
  scale_y_continuous(name = "standardized Weight", (limits = c(min(RATSL$stdWeight), max(RATSL$stdWeight))))
```

The graph drawn with standardized weights shows quite clearly how the weights of the rats are placed in relation to the mean, but also, the relative development of individuals and groups. In group 1, the development seem to have been fairly even, the change is hardly visible. In group 2, individuals move in opposite directions. In group 3, weigths seem to be about steady or declining.

Caution should be taken in drawing far-reaching conclusions from this data, especially considering the small number of individuals included in groups 2 and 3.

Let's move on to summary graphs.
```{r}
# Summary data with mean and standard error of RATS by treatment and week 
RATSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise(mean = mean(Weight), se = (sd(Weight)/sqrt(length(unique(RATSL$ID))))) %>%
  ungroup()
# Glimpsing the data
glimpse(RATSS)
# Plotting the mean profiles
ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.9,0.4)) +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)")
```

Basically, the graph above shows that the variance in groups 1 and 3 is very small, and only slightly bigger in group 2. From the graph it would be roughly discernible that the weight is rising faster in group 2 than in groups 1 and 3.

We can also create boxplots in order to observe outliers and identify thresholds for filtration of outliers.

```{r}
# Create a summary data by Group and ID with mean as the summary variable (ignoring baseline Time 1).
RATSL8S <- RATSL %>%
  filter(Time > 1) %>%
  group_by(Group, ID) %>%
  summarise(mean=mean(Weight) ) %>%
  ungroup()

# Glimpsing the data
glimpse(RATSL8S)

# Drawing a boxplot of mean versus Group
ggplot(RATSL8S, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), 64 days")
```

If we are to follow the recipe of the Datacamp course and MABS book precisely, I would as my next step filter out the outlier values in order to plot the data again. I can do so by cutting off values below 250 grams and above 550 grams. However, getting rid of the outlier of group 3 will be trickier - I might have to filter out rats under 500 grams in that group only.

Again, I want to stress the small sample size, since this poses considerable problems for drawing conclusions. I would even be willing to pose some questions relating to the design of the original study Should for instance groups 2 and 3 have been combined, since the rats in these groups are considerably larger than in group 1 and since the groups taken together would be equally large as group 1? Or should the groups have been mixed, so that an equal number of small and large rats would have been given the same diet?

```{r}
# Filtering out outliers (rats weighting more than 550 and less than 250 grams)
RATSL8S1 <- RATSL8S %>% filter(RATSL8S$mean<550 & RATSL8S$mean > 250)
# Filtering the last outlier (in group 3, weighting less than 500 grams)
RATSL8S1 <- RATSL8S1[-which(RATSL8S1$Group == 3 & RATSL8S1$mean < 500),]
# Inspecting the summary reveals that one outlier has been eliminated in every group
summary(RATSL8S1)
# Plotting the data again without outliers
ggplot(RATSL8S1, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), 64 days")
# Inspecting summaries by group
RATSL8S1 %>% 
  group_by(Group) %>% 
  summarize(mean = mean(mean))
```

The plot drawn without outliers confirms largely what we knew from before: that group 1 consists of small rats, group 2 of mid-size and group 3 of large rats, weighting 269, 452 and 538 grams on average. The variance within groups is very small, partly due to the small sample size and the subsequent filtration of outliers.

Let's do some calculations. Since I have three groups, I will not be performing simple t-tests between the groups since this could lead to the accumulation of type I errors. Instead, I'll rely on ANOVA for testing. 

The analysis confirms that the baseline value for Weight is strongly related to the Weight values taken on day 1 (p < 0.001). There is some evidence for the diet given to Group 2 being efficient (p = 0.07586). Printing a regression summary of the fit confirms that the diet given to Group 3 has no statistically significant impact on the weights of rats.

```{r}
# Adding the baseline. According to the instructions on the Data camp course site, 
# we should be using the original data as the source for the new variable. 
# Since I never imported the original data, I now have to extract the baseline from 
# the imported data. (I did confirm that the extracted vector corresponds to that from the original data.)
RATSL8S2 <- RATSL8S %>%
  mutate(baseline = RATSL[RATSL$Time == 1,]$Weight)

# Fit the linear model with the mean as the response 
fit <- lm(mean ~ baseline + Group, data = RATSL8S2)

# A command used to prompt the use of zeros instead of scientific notation for p values
options(scipen=999)

# Compute the analysis of variance table for the fitted model with anova()
anova(fit)

# Print a summary of the regression model for comparison.
summary(fit)
```

The confidence intervals confirm that 95 per cent of rats in Group 2 gained up to 76 grams of weight or lost at the most 6 grams, when comparing their mean weights with the baseline. About as large weight gains were reported in groups 1 (visible as the Intercept) and 3 (up to 80 grams when comparing mean weights with baseline). However, in these groups there was a greater possibility of losing weight during the observation period.

```{r}
# The confidence intervals (at 95 %) are given below.
confint(fit)
```




### Exploring the BPRS dataset

In the second part of the exercise we will be exploring the BPRS dataset using the techniques employed in Chapter 9 of the course book. Vehkalahti and Everitt (2019: 157) summarize the psychiatric study that provided the data for the dataset: "40 male subjects were randomly assigned to one of two treatment groups and each subject was rated on the brief psychiatric rating scale (BPRS) measured before treatment began (week 0) and then at weekly intervals for eight weeks. The BPRS assesses the level of 18 symptom constructs such as hostility, suspiciousness, hallucinations and grandiosity; each of these is rated from one (not present) to seven (extremely severe). The scale is used to evaluate patients suspected of having schizophrenia."

Again, we will begin by reading the data, converting categorical variables to factors, exploring the structure of the data and plotting the data in order to explore patterns.

```{r}
# Reading the data
BPRSL <- read.csv("./data/bprsl.csv", stringsAsFactors = T)
# Converting categorical variables to factors
BPRSL$treatment <- as.factor(BPRSL$treatment)
BPRSL$subject <- as.factor(BPRSL$subject)
# Confirming the structure of the dataset
str(BPRSL)
# Viewing the data
head(BPRSL)
# Plotting the BPRSL data
ggplot(BPRSL, aes(x = week, y = bprs, group = interaction(treatment, subject), inherit.aes = FALSE, colour = treatment)) + 
  geom_line() + 
  scale_x_continuous(name = "Weeks") + 
  scale_y_continuous(name = "BPRS score") + 
  theme(legend.position = c(0.85,0.85)) + 
  ggtitle("Impact of treatment on BPRS") + 
  theme(plot.title = element_text(hjust = 0.5))
```

From the plot above, it appears as if both treatments have a positive impact, generally lowering BPRS scores. However, simply by casting an eye on the chart it is hard to tell which treatment does a better job.

As a means of initial exploration, we can fit a linear regression model to try to understand the difference between the two models.

```{r}
# creating a regression model
BPRS_reg <- lm(bprs ~ week + treatment, data=BPRSL)
# printing out a summary of the model
summary(BPRS_reg)
```

From the summary of the regression model it can be seen that time (variable week) has a clearly significant impact. That is, the more weeks that pass, the lower the bprs score. However, we cannot find any difference between treatments, at least not when inspecting results using the regression model ignoring the repeated-measures structure.

Let's move on to fit a random intercept model, which will allow the linear regression fit for each individual to differ in intercept from other individuals. We use the same explanatory variables: time (week) and treatment.

We will use the lme4 package, as instructed. The formula adheres to the basic linear regression formula standards, with the addition of random-effects terms distinguished by vertical bars (|).

```{r}
# accessing the lme4 library
library(lme4)
# creating a random intercept model
BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)
# Printing the summary of the model
summary(BPRS_ref)
```

The printout contains a large number of interesting facts. Perhaps the most significant which I would like to pay attention to is the variability of the subjects' scores, expressed by the standard deviation of 6.885 scores.

In the following, we can fit a random intercept and random slope model to the bprs data. This will allow the linear regression fits for each individual to differ in intercept and in slope, which will make it possible to account for the individual differences in the subjects' developments and the effect of time.

```{r}
# creating a random intercept and random slope model
BPRS_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRSL, REML = FALSE)
# printing a summary of the model
summary(BPRS_ref1)
# perform an ANOVA test on the two models
anova(BPRS_ref1, BPRS_ref)
```

According to the printout of the random intercept and random slope model, the individual standard deviation has grown to over 8 now. The ANOVA test of the two models suggests that the random intercept and random slope model (BPRS_ref1) makes a better fit against the random intercept model (p = 0.026). Thus, the individual slopes seem to have an impact.

We can now fit a random intercept and slope model that allows for interaction between time and treatment.

```{r}
# creating a random intercept and random slope model with an interaction term
BPRS_ref2 <- lmer(bprs ~ week + treatment + (week | subject) + week*treatment, data = BPRSL, REML = FALSE)
# printing out a summary of the model
summary(BPRS_ref2)
# performing an ANOVA test on the two models
anova(BPRS_ref2, BPRS_ref1)
```

From the summary, it does not appear as if the interaction term would have any significant effect on the model as a whole (t value 1.785). Neither does treatment have any statistically significant impact (t value -1.2). 

As a final step, we can reprint the original observed values study how they compare with the fitted values of the model. As it appears, the trend visible in the observed values is even more clearly visible in the model using the fitted values. I would conclude that time spent in care clearly has an effect on the brief psychiatric rating scale score measured among the 40 men here. Whether the decline is *due* to the treatment is unclear. From the data analysed here, it cannot be concluded that one treatment would be better than the other, or that the treatment would be effective in the first place. For us to be able to advance such a claim, we would have needed a control group receiving no treatment at all.

```{r}
# Creating a vector of the fitted values
fitted <- fitted(BPRS_ref2)
# Create a new column fitted to BPRSL
BPRSL <- mutate(BPRSL, fitted=fitted)
# Reprinting the plot with observed BPRS values (for the sake of comparison)
ggplot(BPRSL, aes(x = week, y = bprs, group = interaction(treatment, subject), colour = treatment)) + 
  geom_line() +
  scale_x_continuous(name = "Time (weeks)") +
  scale_y_continuous(name = "BPRS score") +
  theme(legend.position = c(0.85,0.85)) + 
  ggtitle("Observed BPRS values") + 
  theme(plot.title = element_text(hjust = 0.5))
# plotting the fitted values
ggplot(BPRSL, aes(x = week, y = fitted, group = interaction(treatment, subject), colour = treatment)) +
  geom_line() +
  scale_x_continuous(name = "Time (weeks)") +
  scale_y_continuous(name = "BPRS score") +
  theme(legend.position = c(0.85,0.85)) + 
  ggtitle("Fitted BPRS values") + 
  theme(plot.title = element_text(hjust = 0.5))
```

